---
title: "Seeds of Success: Collection Diversity, Richness, and Conservation "
author: Colin Khoury, Stephanie Green, Daniel Carver 
date: June 22, 2018 
output:
  html_document:
    theme: 'yeti'
    highlight: 'tango'
    toc: true
    code_folding: hide
    df_print: paged
    toc_float:
      collapsed: false
      smooth_scroll: false
---

# Goal 

Aim of article- examine and report on the diversity collected via the BLM SOS program, and its past as well as potential future uses

# Introduction 

Total diversity collected (species, # of samples, maybe make maps of spatial distribution of collected diversity)
How many are CWR

```{r, echo = FALSE}
#import SOS data 
sosData <- read.csv("H:\\SOS_Project\\analysisData\\SOSdata_20180614_forARS.csv")

#Find the total number of recorded entries 
totalSamples <- nrow(sosData)
print(totalSamples)
```

```{r, echo=FALSE}
#Determine the total number of unique families collect and the number of collection in each family 
uniqueFam <- nrow(unique(sosData["FAMILY"]))

## Determine the total number of unique GENUS collect and the number of collection in each GENUS 
uniqueGen <- nrow(unique(sosData["GENUS"]))
print(uniqueGen)


## Determine the total number of unique SPECIES collect and the number of collection in each SPECIES 
uniqueSpe <- nrow(unique(sosData["SPECIES"]))

## Determine the total number of unique SubSPECIES collect and the number of collection in each SubSPECIES 
subSpecies <- sosData[grep("ssp." ,sosData$NAME),]# for some reason the ['colName'] structure does not seem
# work with the grep function
uniqueSub<- nrow(unique(subSpecies["NAME"]))

## Determine the total number of unique varieties collected and the number of collections in each varieties 
varSpecies <- sosData[grep("var." ,sosData$NAME),] # for some reason the ['colName'] structure does not seem
# work with the grep function
uniqueVar<- nrow(unique(varSpecies["NAME"]))

# rather then a print statement I think this should be a nice table. All outputs should be something interesting too look at. 
print(paste( "This data set contains a total of "
             ,totalSamples, "samples. Of which "
             ,uniqueFam," plant families, "
             , uniqueGen, " plant genus and "
             ,uniqueSpe, " unique species are contained" ) )
```

```{r, echo = FALSE}
# histogram with added parameters
collectionYears <- sosData$COLL_YR
hist(collectionYears,
     main="Accessions collected by SOS",
     xlab="Collection Year",
     xlim = c(2000,2018),
     ylab="Relative Abundance",
     col="darkmagenta"
)

```




# Conservation 

- How much adopted by Curators, how much "orphaned" at W6 - need to pull from GRIN Global (examine "site") 
- How much is at NLGRP for long term backup
- Viability/conservation issues (Annette) - proportion easy to germinate vs not, other storage issues


# Uses

Report on distributions out of NPGS of materials from SOS, by year, by receiver, by species or taxonomic group
Investigate reasons for use- narrative description in GRIN Global
How many are CWR and impacts

# Discussion

- W6 doesnt have the resources to manage seeds indefinitely
- Constraints / challenges to adopting more wild plants
- Constraints with regard to getting materials into long term storage
- Challenge of not having seed testing techniques/rules worked out for species


## Quarterly Results {.tabset .tabset-fade .tabset-pills}


```{r, echo = FALSE}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).
