---
title: 'Seeds of Success: Collection Diversity, Richness, and Conservation '
author: "Colin Khoury, Stephanie Green, Daniel Carver"
date: "June 22, 2018"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: yeti
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#import all necessary libraries 
#install.packages("dplyr", "lemon", "ggplot2", "plyr",'tidyr' #Keep this commented out to knit: look into how to get around this. 
library(dplyr)
library(lemon)
library(ggplot2)
library(plyr)
library(tidyr)
#install.packages('devtools')
#library(devtools)
#install_version("rmarkdown",version=1.8)
```

```{r ref.label='lemon' ,echo=FALSE}
#this allow me to define code chunks can call them at later times. Kinda tricky not so sure of the applicaiton just yet/. 
```



# Goal 

Summarize the relationship between the BLM Seeds Of Success(SOS) program and the USDA ARS division. These two organizations are leading the nation in collecting, conserving, and preserving the native plants present on public lands. 

# Introduction{#Introduction}
```{r, message=FALSE, warning=FALSE, echo=FALSE}
#import SOS data 
sosData <- read.csv("H:\\SOS_Project\\analysisData\\SOSdata_20180614_forARS.csv")

#Find the total number of recorded entries 
totalSamples <- nrow(sosData)
```

### Why SOS program started, what was aims, timeline, etc
The Seeds of Success program was established in 2001 to facilitate the collection of native seeds. To make use of the seeds collected, the SOS program partners with land management agencies to support the re-vegetation of areas, academic and professional partners to promote and advance scientific research, and the United States Department of Agriculture for the long term preservation of species genetic diversity.    

 - fill this out to capture aims
 
 - continue to add content to this as I learn more; ask Colin and Stephanie to add to this background 


```{r, caption="Number of Collections per Family from the SOS program." }
# histogram with added parameters
sosData <- read.csv("H:\\SOS_Project\\analysisData\\SOSdata_20180614_forARS.csv")
glimpse(sosData)

# collectionYears <- sosData$COLL_YR
# p1 <- hist(collectionYears,
#   main="Accessions collected by SOS",
#   xlab="Collection Year",
#   xlim = c(2000,2018),
#   ylim = c(0,3500),
#   ylab="Number of Accession Collected",
#   border="white",
#   col="blue"
# )

ggplot(sosData, aes(COLL_YR))+
  geom_histogram(binwidth = 1, 
                 col="white",
                 fill="blue",
                 alpha = 0.5) + 
  labs(title = "Number of Collections")+
  labs(x = "Collection Year", y= "Number of Accessions") + 
  xlim(c(2000,2018))+
  ylim(c(0,3500))+
  scale_x_continuous(breaks = seq(2000,2018, 2))+
  scale_y_continuous(breaks = seq(0,3500, 500))+
  theme_minimal()

```


### But- not well known how much diversity has been collected in total, and what its uses have been
Detailed collection methodologies mean that SOS accessions have information about the species collected, the region it was collected from, the land management agency where that collection occurred, and positional data in the form of latitude, longitude, and elevation. This data provides information about the seeds collected but does not tell us anything about what happened to the seeds after the collection occurs. The article is meant to create those connections and tell the full story of how these seeds are providing services to the landscape and understanding of the species. 


### Therefore- article examining these issues
This will be done by accessing how many of these collections represent economically useful and/or  crop wild relatives at both the national and international scale. The relationship between the SOS collections and the ARS collection will be examined to determine how many collections are store in long term storage and who is using the seeds outside of the ARS and for what purpose. 

- this can be added to. It's the first time we talk about CWR and there need more of an introductions Specific the two datasets used 


### Total diversity collected (
- just charts for the momments. I want to generate a map for this section
```{r, echo=FALSE, tidy=TRUE, message=FALSE, warning=FALSE}
#Determine the total number of unique families collect and the number of collection in each family 
uniqueFam <- dplyr::summarise(sosData,n_distinct(sosData$FAMILY))
uniqueFam1 <- dplyr::count(sosData,FAMILY) #keep this in hear incase we want to display it 

## Determine the total number of unique GENUS collect and the number of collection in each GENUS 
uniqueGen <-  dplyr::summarise(sosData,n_distinct(sosData$GENUS))
uniqueGen1 <- dplyr::count(sosData, GENUS)

## Determine the total number of unique SPECIES collect and the number of collection in each SPECIES 
uniqueSpe <-  dplyr::summarise(sosData,n_distinct(sosData$SPECIES))
uniqueSpe1 <- dplyr::count(sosData, SPECIES)


## Determine the total number of unique SubSPECIES collect and the number of collection in each SubSPECIES 
subSpecies <- sosData[grep("ssp." ,sosData$NAME),]# for some reason the ['colName'] structure does not seem
# work with the grep function
uniqueSub<- dplyr::summarise(subSpecies,n_distinct(subSpecies$NAME))

## Determine the total number of unique varieties collected and the number of collections in each varieties 
varSpecies <- sosData[grep("var." ,sosData$NAME),] # for some reason the ['colName'] structure does not seem
# work with the grep function
uniqueVar<- dplyr::summarise(varSpecies,n_distinct(varSpecies$NAME))

## DEfeine row and column names 
labelsRow1 <- c("SOS Collection")
labelsCol1 <- c('All Accessions', "Total Number of Familes", "Total Number of Genus", "Total Number of Sub Species")

#Initiate an empty dataframe 
sosSummary <- data.frame()

# create the first row of the dataframe 
sosData <- c(totalSamples, uniqueFam, uniqueGen, uniqueSpe)

#bind the two df together 
sosSummary <- tbl_df(rbind(sosSummary, sosData))
#define column names and row names 
colnames(sosSummary) <- labelsCol1
rownames(sosSummary) <- labelsRow1
sosSummary

```




This Seeds of Success program has collected a total of `r totalSamples` seed samples. Of which `r uniqueFam[1,]` plant families
`r uniqueGen[1,]`  plant genus and `r uniqueSpe[1,]` unique sub species are represented. 

- this area intesting numbers but some knowledge on how they relate to the total number of species present in the country might make the factiods stand out some more


```{r caption="Number of Collections per Family from the SOS program.", rows.print=10}
families <- dplyr::arrange(uniqueFam1, desc(n)) 
colnames(families) <- c("Family", "Total Number of Accessions")
families

```
An accession is a collection of seeds from a single sampling location. There are enough seeds in each accession to split the collection for short term storage, long term preservation, and distribution. Out of the `r uniqueFam[1,]` familes collected half of the total accessions collected come from just 5 families. The commonality and number of species represented by the top 5 collected families may explain the observed relationship. 


### How many are Crop Wild Relatives or Economic Plants

The importance of this collection is increased by the fact that many of the seed taxa that were collected by the SOS program are defined as either Crop Wild Relatives or World Economic Plants. Crop Wild Relatives (CWR) are the genetic predecessors of modern crops and are found throughout the diverse ecological and bio physiological environments of the world. The variability in growing conditions and potential geographic isolation between stands has shown that WCRs contain a higher level of genetic diversity (*reference this*) . The subsequent traits found within WCR are used to add to the resistance, quality, and breeding efficiency of modern crops (*reference this*).

- add a bit about the various economic use categories define by GRIN 

- talk about the WEP data set; how it was developed and why we are using it 
*get reference for the WEP dataset received from Daniel*.

- expand upon the k13 data source a bit to explain why we are using it.  


```{r caption="Summary of the SOS collection relative to Econmically Useful and Wild Crop Relatives.", rows.print=15}

###
# Create a summary chart showing the total number of accession, total taxa, # of WCR NA, # eco NA, # of economic WEP  
# this may be added to also include the data about what's managed by the usda and what's at the NLGRP 
#

# Replace the name column of the sos data so that the "ssp." notation is replaced with "subsp." as defined
# by the grin Data base 

#import SOS data 
sosData <- read.csv("H:\\SOS_Project\\analysisData\\SOSdata_20180614_forARS.csv")
# created a new column that has the same col name as the K13 data set. I will use the col to join on 
sosData$Taxon <- gsub("ssp.","subsp.", sosData$NAME) 

#detemine the total number of taxon in dataset 
sosTotalUnique <- nrow(unique(sosData[1]))

# read in the K13 data 
K13Data <- read.csv("H:\\SOS_Project\\analysisData\\CWR_US_inventory_2013_7_10.csv")

#detemine the total number of taxon in dataset 
K13TotalUnique <- n_distinct(K13Data$Taxon)

# join the data sets
combinedData <- join(sosData, K13Data, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the K13 specific data columns are dropped. 
trimCombinedData <- filter(combinedData, !is.na(Taxon_w_author))

#subset all WIld Utilizes Species and calculater total number of species 
wusSamples <- filter(trimCombinedData, Type== 'WUS')
uniqueWUS <- n_distinct(wusSamples$Taxon)

#subset all Crop Wild RElatives and calculater total number of species 
cwrSamples <- filter(trimCombinedData, Type== 'CWR')
uniqueCWR <- n_distinct(cwrSamples$Taxon)

### 
#Read in the WEP dataset 
# read in the WEP data 
WEPData <- read.csv("H:\\SOS_Project\\analysisData\\EconomicUseCategoriesOfSpecies.csv")

## Rename the name column to match the newly created column from the SOS data 
colnames(WEPData)[2] <- "Taxon"

#detemine the total number of taxon in dataset 
WEPTotalUnique <- n_distinct(WEPData$Taxon)

# join the data sets
combinedData2 <- join(sosData, WEPData, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the WEP specific data columns are dropped. 
trimCombinedData2 <- filter(combinedData2, !is.na(economic_usage_code))

#calculate total number of species from 
uniqueWEP <- n_distinct(trimCombinedData2$Taxon)

## DEfeine row and column names 
labelsRow <- c("SOS Collection")
labelsCol <- c('All Accession', "Total Number of Taxon", "Total Number of WUS from K13", "Total Number of CWR from K13", "Total Number of WEP")

#Initiate an empty dataframe 
sosSummary2 <- data.frame()

# create the first row of the dataframe 
sosCollection <- c(totalSamples, sosTotalUnique, uniqueWUS, uniqueCWR, uniqueWEP)

#bind the two df together 
sosSummary2 <- rbind(sosSummary2, sosCollection)
#define column names and row names 
colnames(sosSummary2) <- labelsCol
rownames(sosSummary2) <- labelsRow
sosSummary2


```
This chart compared the SOS data against two data sets that have define Wild Utilized Species(WUS), Crop Wild Relatives(CWR), and World Economic Plants(WEP) at the taxa level. This illustrates that at least a 1/6 or of the taxa collected by SOS are considered to be useful plants outside of thier ecological value.  



# Conservation 

The Seeds of Success program provides the boots on the ground effort in the collection of seed resources. The program partners with the Agricultural Research Service of the USDA to store, maintain, and distribute the seeds collected. The Pullman, Washington section of the ARS is the group responsible for the initial on boarding and entering of the seeds into the Genetic Resource Inventory Network (GRIN) as well as maintaining the seeds in environments that support preservation.  

### stored within ARS system 
```{r caption="Summary of SOS and GRIN Relative to CWR and WEP", rows.print=15}
# read in GRIN Accessions 
grinData <- read.csv("H:\\SOS_Project\\analysisData\\sosAllAccessions.csv")

# calcualte the total number of features and the total number of taxon 
totalSamplesGRIN <- nrow(grinData)# why would these two numbers be different? Question for Renee 
totalTaxon <- dplyr::n_distinct(grinData$Taxon)

# join the the GRIN data with the K13 Dataset 
combinedData3 <- join(grinData, K13Data, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the K13 specific data columns are dropped. 
trimCombinedData3 <- filter(combinedData3, !is.na(Taxon_w_author))

#subset all WIld Utilizes Species and calculater total number of species 
wusSamples3 <- filter(trimCombinedData3, Type== 'WUS')
uniqueWUS3 <- n_distinct(wusSamples3$Taxon)

#subset all Crop Wild RElatives and calculater total number of species 
cwrSamples3 <- filter(trimCombinedData3, Type== 'CWR')
uniqueCWR3 <- n_distinct(cwrSamples3$Taxon)

#### Repeat with the WEP DATAset 
# join the grin data with the WEP data
combinedData4 <- join(grinData, WEPData, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the WEP specific data columns are dropped. 
trimCombinedData4 <- filter(combinedData4, !is.na(economic_usage_code))


#calculate total number of species from 
uniqueWEP4 <- n_distinct(trimCombinedData4$Taxon)

# create the first row of the dataframe 
grinCollection <- c(totalSamplesGRIN, totalTaxon, uniqueWUS3, uniqueCWR3, uniqueWEP4)

#bind the two df together 
sosSummary2 <- rbind(sosSummary2, grinCollection)
rownames(sosSummary2)[2] <- "GRIN Collection"
sosSummary2

```

The difference in the total number of accessions collected by SOS and the number of accessions recorded in  GRIN is significant. Part of this can be explained by a potential backlog of samples within the ARS and a lag in delivery time between collection and distribution. It generally takes a year from when the seeds are collect to when they are received by the ARS.

```{r}
# filter based on year collected and get a count on the number of unique accessions
sosPre2017 <- filter(sosData, COLL_YR < 2017) %>% n_distinct()
percentAtARS = signif(100*(totalSamplesGRIN/sosPre2017),4)
```
For example, a total of `r sosPre2017` accessions were brought into the SOS collection before 2017. The ARS is still processing SOS data from 2017. Therefore, `r percentAtARS` percent of the samples collected by SOS have been brought into the ARS and GRIN Global system. 

- I'm not sure given the data sets that I have I'll be able to join the two features. All the SOS ID are lost in the grin query.

- this is an SOS-GRIN-UNION sheet that was send by the SOS, I'm just exploring here for a bit 

Using a dataset generated by the BLM which joins all SOS data and all ARS data we were able to acess difference across the two data records. 

```{r}
# read in union data 
sosUnion <- read.csv("H:\\SOS_Project\\analysisData\\SOS_ARSUNION_20180430.csv")
#generate date showing presences comments 
sosUnion1 <- count(sosUnion$RH_COMMENTS) %>%
  dplyr::rename(Condition = x, NumberOfAccession = freq )
#print tbl
sosUnion1

```

- this is a union that was performed by the SOS program. I've not been able to reproduce the process via GRIN: probably could with help, but 

**interesting points**

1.  10,000 samples from SOS are not in GRIN or ARS. ~ 2000 of these were sampled in 2017 or later and that could explain why the two data sets are not connecting 

2. 727 samples are entered into GRIN as SOS samples but are not present on BLM data

**again this is exploratory** I did not generate the join between the ARS AND SOS data sets to produce these comparisons but the data set brings up some interesting questions on the process. 
I bring this up again because it may be difficult to reproduce this dataset sence the individual who generated it has left the BLM 

```{r}
# filter based on collected by grin but not present 
sosunion2 <- filter(sosUnion, sosUnion$RH_COMMENTS == ("IN SOS_ NOT IN ARS OR GRIN"))

collectionYears2 <- sosunion2$COLL_YR
p2 <- hist(collectionYears2,
  main="Accessions collected by SOS that are not in ARS or GRIN ",
  xlab="Collection Year",
  xlim = c(2000,2018),
  ylim = c(0,3500),
  ylab="Number of Accession Collected",
  border="white",
  col="dark green"
)

ggplot(sosunion2, aes(COLL_YR))+
  geom_histogram(binwidth = 1, 
                 col="white",
                 fill="dark green",
                 alpha = 1) + 
  labs(title = "Accessions collected by SOS that are not in ARS or GRIN", 
       x = "Collection Year",
       y= "Number of Accession Collected",
       caption = "") + 
  xlim(c(2000,2018))+
  ylim(c(0,3500))+
  scale_x_continuous(breaks = seq(2000,2018, 2))+
  scale_y_continuous(breaks = seq(0,3500, 500))+
  theme_minimal()+
  geom_histogram(data= sosData, 
                 aes(sosData$COLL_YR),
                 binwidth = 1,
                 col="white",
                 fill= "grey",
                 alpha = 0.5)
  

#check out the site for more [information](https://stackoverflow.com/questions/6957549/overlaying-histograms-with-ggplot2-in-r)

```

If you refer back to the [Introduction](# Introduction) you can see that the there is a higher proportion of seeds being collected in 2015 and 2016 that are not making it into the ARS system. This may reflect a backlog in process. 

### How much is at NLGRP for long term backup

Once within the ARS system the quality of the seed collections are accessed and samples that are deemed beneficial to preserve over an extended period of time are transfer to the National Laboratory for Genetic Resource Preservation in Fort Collins, Colorado. The NLRGP specializes in the long term storage of genetic resources. If a set of seeds is stored at NLGRP it can be assured that it is protect and preserved to the highest standard. The coordinating efforts to bring seeds into the long term storage is essential to the longevity of the genetic material store within the seeds. 

```{r}
#filter based on the "is backuped" column.  
backedUp <- dplyr::filter(grinData, Is.Backed.Up. == 'Y' )
totaledBackedUp <-nrow(backedUp)
#filter based on the backup location being NLGRP 
nlgrpData <- dplyr::filter(backedUp, Backup.Location.1 == 'NSSL' )
totalBackedUp <- nrow(nlgrpData)
## Example of the backed up features that are missing a backup location????
#### does this mean an error in data entry, can we assume 
MissingBackup <- count(backedUp$Backup.Location.1 )

#define the total number of species present in backup
totalTaxonNLGRP <- n_distinct(nlgrpData$Taxon)

### apply simliar subsets based on the connection between K13 dataset
# join the the GRIN data with the K13 Dataset 
combinedData5 <- join(nlgrpData, K13Data, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the K13 specific data columns are dropped. 
trimCombinedData5 <- filter(combinedData5, !is.na(Taxon_w_author))

#subset all WIld Utilizes Species and calculater total number of species 
wusSamples5 <- filter(trimCombinedData5, Type== 'WUS')
uniqueWUS5 <- n_distinct(wusSamples5$Taxon)

#subset all Crop Wild RElatives and calculater total number of species 
cwrSamples5 <- filter(trimCombinedData5, Type== 'CWR')
uniqueCWR5 <- n_distinct(cwrSamples5$Taxon)


#### Repeat with the WEP DATAset 
# join the grin data with the WEP data
combinedData6 <- join(nlgrpData, WEPData, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the WEP specific data columns are dropped. 
trimCombinedData6 <- filter(combinedData6, !is.na(economic_usage_code))

#calculate total number of species from 
uniqueWEP6 <- n_distinct(trimCombinedData6$Taxon)

# create the first row of the dataframe 
nlgrpCollection <- c(totalBackedUp, totalTaxonNLGRP, uniqueWUS5, uniqueCWR5, uniqueWEP6)

#bind the two df together 
sosSummary2 <- rbind(sosSummary2, nlgrpCollection)
rownames(sosSummary2)[3] <- "NLGRP Collection"
sosSummary2


```



### How much adopted by Curators, how much "orphaned" at W6 - need to pull from GRIN Global (examine "site") #Not sure how to answer this some at this time 

The challenge of coordinating the preservation of these seeds is seen in part by how my samples become orphaned within the system. Orphaned seeds are those which do not fit into any of the existing collections. As a results seed curators need to assess how the seeds can best be integrated and is it worth incorporating that sample. While this decision making process is occurring the seeds are not being backed up and are not grown out to plants to replenish the existing seed population. 

```{r}
# call in sosAccessions: summaries by maintaince site or owned by ***check with Colin or Renee on this one***
# produce a table showing number of successions and proportion of total

sosAccessions <- read.csv("H:\\SOS_Project\\analysisData\\sosAllAccessions.csv")

# create a count list based on the number of values within the tbl
sosAcc <- mutate(sosAccessions, count =1 )

# subet columns needed.
sosOwner <- count(sosAcc$Accession.Prefix)%>%
  arrange(desc(freq))%>%
  dplyr::rename(Facility = x, NumberOfAccessions = freq)


sosOwner
```
The majority of the SOS collection is maintained by the ARS in Pullman, Washington (W6). This is where the native plant collection is based. Accession that are maintained at other locations are there due to the request of other curators. The are accession that are very unlikely to be orphaned. It is diffent to determine from data alone how many of the accession at W6 an being actively store but not regenerated. 


### Viability/conservation issues (Annette) - proportion easy to germinate vs not, other storage issues

Collection and storage are only part of the story. Germination and the ability to grow out the plant to collect new seeds is a large part of what the ARS supports. The troublesome fact about the SOS seeds is that these species have not often been collected and maintained in seed banks before the initiation of the SOS program. Often little is known about the conditions required to germinate the seeds. 
```{r}
viability <- read.csv("H:\\SOS_Project\\analysisData\\generaAverageGermination.csv")
viability <- tbl_df(viability)
uniqueGenus <- n_distinct(viability$Genus)
percentGenus <- signif(100*(uniqueGenus/uniqueGen),4)
```

An experiment was developed and performed by Annette *find her last name* of NLGRP. Annette tested the viability of seeds to germinate for `r uniqueGenus` unique genus. This accounts for `r percentGenus[1,]` percent of all the genus collected by the SOS program. The data generated by this study illustrates what to expect when working with these plants in the future. 


```{r caption="Listing of Seed Viability by Genus", rows.print=10}
library(dplyr)
viability <- read.csv("H:\\SOS_Project\\analysisData\\generaAverageGermination.csv")
viability <- tbl_df(viability)
colnames(viability) <- c('Genus', 'percentViable' )
arrange(viability, desc(percentViable))

```

```{r caption="Summary of SOS and GRIN Relative to CWR and WEP", rows.print=15}
# histogram with added parameters
hist(viability$percentViable,
  main="Successfully Sprouted Seeds",
  xlab="Percent Successful",
  ylab="Number of Species Tested",
  col="light green",
  border="white", 
  las=1, #defien the orientation of the label 
  breaks=20
)

```


# Uses

### Report on distributions out of NPGS of materials from SOS, by year, by receiver, by species or taxonomic group

The ARS systems stores and distributes the seeds that are collected by SOS. Seeds are distributed for a number of reasons including academic research studies and educational demonstrations. Many of the distributions take place within the ARS when seeds are sent to and from various seed labs. The distribution of seeds is end user driven and a a result the volume of seeds transferred fluctuates over time. 

```{r warning=FALSE, caption="Distribution of seeds over time", rows.print=15}
activeOrders <- read.csv("H:\\SOS_Project\\analysisData\\sosActiveOrders.csv")
date <- separate(activeOrders, Ordered.Date, into = c("month","day", "year"), sep = "/")
# histogram with added parameters
hist(as.numeric(date$year),
  main="SOS Samples Distributed Over Time",
  xlab="Year",
  ylab="Number of Samples Distributed",
  col="grey",
  border="white", 
  las=1 #defines the orientation of the label 
)

```

The seeds can be used for multiple reasons but by far Research and Education are the top two requests. The first bin on this graph is 2005 and the last is 2017. 

```{r ,warning=FALSE ,rows.print=5}
orderType <- count(activeOrders$Intended.Use)
arrange(orderType, desc(freq)) %>% `colnames<-`(c("Reason for Distribution", "Individual Shipments"))
```
These values are written in by individuals who place the order so while this shows trends there are many unique options that do not fall into a group because of the extra text. 

```{r, warning=FALSE}
#read in feature 
distribution <- read.csv("H:\\SOS_Project\\analysisData\\SOSOrdersTaxa.csv")
#summarize by genus and sum the total number of packets sent per Genus 
genusSum <- distribution %>% 
  group_by(genus_name)%>%
  dplyr::summarise(genusSum = sum(COUNT))%>% 
  arrange(desc(genusSum))%>%
  dplyr::rename(Genus = genus_name  ,IndividualSeedsSent = genusSum)

genusSum

#calculate the total number of seeds sent 
totalSeedsSent <- sum(distribution$COUNT)

```


Table illustrating the total number of seeds sent per genus. There has been a total of `r totalSeedsSent` seeds distributed by the ARS that were collected by the SOS program. 


```{r, warning=FALSE}

K13Data$genus_name <- K13Data$Genus



#Call in the WUP/CWR data, summarize by Genus, join to distribution data by genus 
K13genus <- K13Data %>%
  group_by(genus_name) %>%
  summarise_()%>%
  inner_join(distribution,by="genus_name")

numberGenus <- n_distinct(K13genus$genus_name)

# calculated the total number of seeds sent 
totalWUSSet <- sum(K13genus$COUNT)

#define the unique number of genus and seeds sent to enitity groups 
reciever <- distribution %>%
  group_by(title) %>%
  dplyr::summarise(numberOfGenus = n_distinct(genus_name), numberOfSeeds = sum(COUNT)) %>%
  arrange(desc(numberOfGenus))%>%
  dplyr::rename(EntityRecievingSeeds = 'title'  ,NumberOfGenusDelivered = 'numberOfGenus', IndividualSeedsDelivered = 'numberOfSeeds')  

reciever

```
There are a total of `r totalSeedsSent` seeds that have been distributed by the ARS. Of that `r totalWUSSet`, are considered to be either Wild Utilized Plants or Crop Wild Relatives. This is representing `r numberGenus` unique genus



```{r, warning=FALSE}
# call in data on the distribution of species 
speciesOrdered <- read.csv("H:\\SOS_Project\\analysisData\\sosOrderItems.csv")
speciesOrdered <- mutate(speciesOrdered, Order.Request.ID = as.factor(order_request_id))

# one to main join with activeOrders based on "Order.Request.ID"
speciesOr <- full_join(speciesOrdered, activeOrders, by="Order.Request.ID")


accessionsSent <- arrange(count(speciesOr$accession_number) , desc(freq)) %>% `colnames<-`(c("Accession", "Accessions Pulled From"))
accessionsSent 
```
This chart shows the number of times an order was place for a specific accession. 

```{r, warning=FALSE}
taxaSent <- arrange(count(speciesOr$Taxon) , desc(freq)) %>% `colnames<-`(c("Taxa", "Times Delivered"))
totalSent <- sum(taxaSent$`Taxa Delivered`)
#totalSent -- not super relavent because it is the number of time indiviudal species were sent, 
taxaSent
```
This chart shows the number of times a specific taxa was has been delivered. These deliveries could come from multiple accessions.


```{r, warning=FALSE}
requestor <- arrange(count(speciesOr$Requestor) , desc(freq)) %>% `colnames<-`(c("Entity Recieving Seeds", "Individual Species Sent"))
requestor

```
This graphic shows which location is receiving the greatest diversity of taxon 
```{r}
### come back to this at some point 

# deliveryData <-read.csv("H:\\SOS_Project\\analysisData\\sosOrdersTotal.csv")
# print(deliveryData)
# 
# # select all rows that were delivered to ARS locations 
# ars <- filter(deliveryData, category_code == "UARS")
# ars <- ars[,1:2]
# labelsColumns <- c('Number of Deliveries', "Reason for Delivery" )
# values2 <- c("Replenishment", "Germination", "Observation"
#                               ,"Information only", "Transfer", "Backup"
#                               ,"Distribution")
# ars$order_type_code <- values2
# 
# colnames(ars) <- labelsColumns
# colnames(ars)
# ars<- ars[order(-ars$'Number of Deliveries'),]

```

# to Do 

- Investigate reasons for use- narrative description in GRIN Global
- How many are CWR and impacts

# Discussion

- W6 doesn't have the resources to manage seeds indefinitely
- Constraints / challenges to adopting more wild plants
- Constraints with regard to getting materials into long term storage
- Challenge of not having seed testing techniques/rules worked out for species


## Quarterly Results {.tabset .tabset-fade .tabset-pills}
