---
title: 'Seeds of Success: Collection Diversity, Richness, and Conservation '
author: "Colin Khoury, Stephanie Green, Daniel Carver"
date: "June 22, 2018"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    theme: yeti
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
---

# CheckList 
## Datasets
- SOSdata = 1
- K13
  - Economic Plants =2 
  - wild crop Relatives = 3 
- GRINSOS
  - all inventory = 4
  - deliveries with taxon = 5 
- WEP =6 

## connections 
- 1&2 = Of all the SOS seeds collected how many species are considered to be Economic plants in north america 
- 1&3 = Of all the SOS seeds collected how may species are considiered to be Wild Crop Relatives in North America 
- 4&2 = Of all the SOS seeds maintained by ARS how many species are considered to be Economic plants in north america 
- 4&3 = Of all the SOS seeds maintained by ARS how many species are considered to be Wild Crop Relatives in north america 
- 5 = Deliveries: Who received, Reason, data, taxa shipped 
- 1&6 = 1&2 = Of all the SOS seeds collected how many species are considered to be World Economic plants 
- 4&6 = 1&2 = Of all the SOS seeds maintained by ARS how many species are considered to be World Economic plants 
- 1 = summary of all taxa collected 

# Goal 

Aim of article- examine and report on the diversity collected via the BLM SOS program, and its past as well as potential future uses

# Introduction 

Total diversity collected (species, # of samples, maybe make maps of spatial distribution of collected diversity)
How many are CWR
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#import all necessary libraries 
#install.packages("dplyr", "lemon", "ggplot2", "plyr") #Keep this commented out to knit: look into how to get around this. 
library(dplyr)
library(lemon)
library(ggplot2)
library(plyr)
```

```{r, message=FALSE, warning=FALSE}
#import SOS data 
sosData <- read.csv("H:\\SOS_Project\\analysisData\\SOSdata_20180614_forARS.csv")

#Find the total number of recorded entries 
totalSamples <- nrow(sosData)
```

```{r, echo=FALSE, tidy=TRUE, message=FALSE, warning=FALSE}
#Determine the total number of unique families collect and the number of collection in each family 
uniqueFam <- dplyr::summarise(sosData,n_distinct(sosData$FAMILY))
uniqueFam1 <- dplyr::count(sosData,FAMILY) #keep this in hear incase we want to display it 

## Determine the total number of unique GENUS collect and the number of collection in each GENUS 
uniqueGen <-  dplyr::summarise(sosData,n_distinct(sosData$GENUS))
uniqueGen1 <- dplyr::count(sosData, GENUS)

## Determine the total number of unique SPECIES collect and the number of collection in each SPECIES 
uniqueSpe <-  dplyr::summarise(sosData,n_distinct(sosData$SPECIES))
uniqueSpe1 <- dplyr::count(sosData, SPECIES)


## Determine the total number of unique SubSPECIES collect and the number of collection in each SubSPECIES 
subSpecies <- sosData[grep("ssp." ,sosData$NAME),]# for some reason the ['colName'] structure does not seem
# work with the grep function
uniqueSub<- dplyr::summarise(subSpecies,n_distinct(subSpecies$NAME))

## Determine the total number of unique varieties collected and the number of collections in each varieties 
varSpecies <- sosData[grep("var." ,sosData$NAME),] # for some reason the ['colName'] structure does not seem
# work with the grep function
uniqueVar<- dplyr::summarise(varSpecies,n_distinct(varSpecies$NAME))

# rather then a print statement I think this should be a nice table. All outputs should be something interesting too look at. 
print(paste( "This Seeds of Success program has collected a total of "
             ,totalSamples, " seed samples. Of which "
             ,uniqueFam," plant families, "
             ,uniqueGen, " plant genus and "
             ,uniqueSpe, " unique sub species are represented" ) )

## DEfeine row and column names 
labelsRow1 <- c("SOS Collection")
labelsCol1 <- c('All Accessions', "Total Number of Familes", "Total Number of Genus", "Total Number of Sub Species")

#Initiate an empty dataframe 
sosSummary <- data.frame()

# create the first row of the dataframe 
sosData <- c(totalSamples, uniqueFam, uniqueGen, uniqueSpe)

#bind the two df together 
sosSummary <- rbind(sosSummary, sosData)
#define column names and row names 
colnames(sosSummary) <- labelsCol1
rownames(sosSummary) <- labelsRow1
sosSummary

uniqueFam[1,]

```
```{r lemon, echo=FALSE}
### 
# testing about the nice charts using lemon and kable
# https://cran.r-project.org/web/packages/lemon/vignettes/lemon_print.html
###
library(lemon)
#knit_print.data.frame <- lemon_print
```

This is a better way to make print statements 

This Seeds of Success program has collected a total of `r totalSamples` seed samples. Of which `r uniqueFam[1,]` plant families
`r uniqueGen[1,]`  plant genus and `r uniqueSpe[1,]` unique sub species are represented.

```{r ref.label='lemon' }
#this allow me to define code chunks can call them at later times. Kinda tricky not so sure of the applicaiton just yet/. 
```


```{r caption="Number of Collections per Family from the SOS program.", rows.print=15}
families <- dplyr::arrange(uniqueFam1, desc(n)) 
colnames(families) <- c("Family", "Total Number of Accessions")
families

```
The Seeds of Success program has been collecting seeds since 2000. 

```{r, caption="Number of Collections per Family from the SOS program." }
# histogram with added parameters
sosData <- read.csv("H:\\SOS_Project\\analysisData\\SOSdata_20180614_forARS.csv")

collectionYears <- sosData$COLL_YR
hist(collectionYears,
  main="Accessions collected by SOS",
  xlab="Collection Year",
  xlim = c(2000,2018),
  ylim = c(0,3500),
  ylab="Number of Accession Collected",
  col="blue"
)

```

The importance of this collection increased by the fact that many of the seed taxa that were collected are defined as either Wild Crop Relatives or World Economic Plants. We will use two datasets that have defined economically useful plants, Khoury et al 2013, and *get reference for the WEP dataset received from Daniel*. 


```{r}

# caption="Summary of the SOS collection relative to Econmically Useful and Wild Crop Relatives.", rows.print=15
# Create a summary chart showing the total number of accession, total taxa, # of WCR NA, # eco NA, # of economic WEP  
# this may be added to also include the data about what's managed by the usda and what's at the NLGRP 
#

# Replace the name column of the sos data so that the "ssp." notation is replaced with "subsp." as defined
# by the grin Data base 
# created a new column that has the same col name as the K13 data set. I will use the col to join on 
sosData$Taxon <- gsub("ssp.","subsp.", sosData$NAME) 

#detemine the total number of taxon in dataset 
sosTotalUnique <- nrow(unique(sosData[1]))


# read in the K13 data 
K13Data <- read.csv("H:\\SOS_Project\\analysisData\\CWR_US_inventory_2013_7_10.csv")

#detemine the total number of taxon in dataset 
K13TotalUnique <- nrow(unique(K13Data[1]))

# join the data sets
combinedData <- join(sosData, K13Data, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the K13 specific data columns are dropped. 
trimCombinedData <- combinedData[!is.na(combinedData$Taxon_w_author), ]

#subset all WIld Utilizes Species and calculater total number of species 
wusSamples <- filter(trimCombinedData, Type== 'WUS')
uniqueWUS <- n_distinct(wusSamples$Taxon)

#subset all Crop Wild RElatives and calculater total number of species 
cwrSamples <- filter(trimCombinedData, Type== 'CWR')
uniqueCWR <- n_distinct(cwrSamples$Taxon)

### 
#Read in the WEP dataset 
# read in the WEP data 
WEPData <- read.csv("H:\\SOS_Project\\analysisData\\EconomicUseCategoriesOfSpecies.csv")

## Rename the name column to match the newly created column from the SOS data 
colnames(WEPData)[2] <- "Taxon"


#detemine the total number of taxon in dataset 
WEPTotalUnique <- n_distinct(WEPData$taxonomy_species_id)

# join the data sets
combinedData2 <- join(sosData, WEPData, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the WEP specific data columns are dropped. 
trimCombinedData2 <- filter(combinedData2, !is.na(economic_usage_code))

#calculate total number of species from 
uniqueWEP <- n_distinct(trimCombinedData2$SPECIES)


## DEfeine row and column names 
labelsRow <- c("SOS Collection")
labelsCol <- c('All Accession', "Total Number of Taxon", "Total Number of WUS from K13", "Total Number of CWR from K13", "Total Number of WEP")

#Initiate an empty dataframe 
sosSummary2 <- data.frame()

# create the first row of the dataframe 
sosCollection <- c(totalSamples, sosTotalUnique, uniqueWUS, uniqueCWR, uniqueWEP)

#bind the two df together 
sosSummary2 <- rbind(sosSummary2, sosCollection)
#define column names and row names 
colnames(sosSummary2) <- labelsCol
rownames(sosSummary2) <- labelsRow
sosSummary2


```


# Conservation 

The Seeds of Success program provide the boots on the ground effort in the collection of seed resources and the program partners with the Agricultural Research Service of USDA to store, maintian, and distribute the seeds collected. The Pullman, Washinto
ng section of the ARS is the group primarily responsable for the initial onboarding and entering of the seeds into the Genetic Resoruce Inventory Network(GRIN) as well as mainting the seeds in environments that support preservation. Once within the ARS system the quality of the seed collections are accessed and samples that are deemed benifical to preserve over an extended period of time are transfer to the National Laboratory for Genetic Resource Preservation in Fort Collins, Colorado. The NLRGP specialized in the long term storage of genetic resources. Below is a few diagrams and table that illistrate the relationship between the Seeds of Success program and the ARS. 




- How much adopted by Curators, how much "orphaned" at W6 - need to pull from GRIN Global (examine "site") #Not sure how to answer this some at this time 


- How much is at NLGRP for long term backup
    - * there are 12959 SOS accessions that are maintianed in long term storage backup at NLGRP.* 
- Viability/conservation issues (Annette) - proportion easy to germinate vs not, other storage issues

```{r caption="Summary of SOS and GRIN Relative to CWR and WEP", rows.print=15}
# read in GRIN Accessions 
grinData <- read.csv("H:\\SOS_Project\\analysisData\\sosAccessions.csv")

# calcualte the total number of features and the total number of taxon 
totalSamplesGRIN <- nrow(grinData)# why would these two numbers be different? Question for Renee 
totalName <- dplyr::n_distinct(grinData$Name)# why would these two numbers be different? Question for Renee 
totalTaxon <- totalName <- dplyr::n_distinct(grinData$Taxon)


# join the the GRIN data with the K13 Dataset 
combinedData3 <- join(grinData, K13Data, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the K13 specific data columns are dropped. 
trimCombinedData3 <- combinedData[!is.na(combinedData3$Taxon_w_author), ]

#subset all WIld Utilizes Species and calculater total number of species 
wusSamples3 <- filter(trimCombinedData3, Type== 'WUS')
uniqueWUS3 <- n_distinct(wusSamples3$Taxon)

#subset all Crop Wild RElatives and calculater total number of species 
cwrSamples3 <- filter(trimCombinedData3, Type== 'CWR')
uniqueCWR3 <- n_distinct(cwrSamples3$Taxon)

#### Repeat with the WEP DATAset 
# join the grin data with the WEP data
combinedData4 <- join(grinData, WEPData, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the WEP specific data columns are dropped. 
trimCombinedData4 <- filter(combinedData4, !is.na(economic_usage_code))

#calculate total number of species from 
uniqueWEP4 <- n_distinct(trimCombinedData4$Taxon)


# create the first row of the dataframe 
grinCollection <- c(totalSamplesGRIN, totalTaxon, uniqueWUS3, uniqueCWR3, uniqueWEP4)

#bind the two df together 
sosSummary2 <- rbind(sosSummary2, grinCollection)
rownames(sosSummary2)[2] <- "GRIN Collection"
sosSummary2

```

```{r}
#filter based on the "is backuped" column.  
backedUp <- dplyr::filter(grinData, Is.Backed.Up. == 'Y' )
totaledBackedUp <-nrow(backedUp)
#filter based on the backup location being NLGRP 
nlgrpData <- dplyr::filter(backedUp, Backup.Location.1 == 'NSSL' )
totalBackedUp <- nrow(nlgrpData)
## Example of the backed up features that are missing a backup location????
#### does this mean an error in data entry, can we assume 
MissingBackup <- count(backedUp$Backup.Location.1 )

#define the total number of species present in backup
totalTaxonNLGRP <- n_distinct(nlgrpData$Taxon)

# apply simliar subsets based on the connection between K13 dataset

# join the the GRIN data with the K13 Dataset 
combinedData5 <- join(nlgrpData, K13Data, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the K13 specific data columns are dropped. 
trimCombinedData5 <- combinedData[!is.na(combinedData5$Taxon_w_author), ]

#subset all WIld Utilizes Species and calculater total number of species 
wusSamples5 <- filter(trimCombinedData5, Type== 'WUS')
uniqueWUS5 <- n_distinct(wusSamples5$Taxon)

#subset all Crop Wild RElatives and calculater total number of species 
cwrSamples5 <- filter(trimCombinedData5, Type== 'CWR')
uniqueCWR5 <- n_distinct(cwrSamples5$Taxon)


#### Repeat with the WEP DATAset 
# join the grin data with the WEP data
combinedData6 <- join(nlgrpData, WEPData, by='Taxon', type='left', match='all')

#trim the data so that taxon that do not have values for the WEP specific data columns are dropped. 
trimCombinedData6 <- filter(combinedData6, !is.na(economic_usage_code))

#calculate total number of species from 
uniqueWEP6 <- n_distinct(trimCombinedData6$Taxon)




# create the first row of the dataframe 
nlgrpCollection <- c(totalBackedUp, totalTaxonNLGRP, uniqueWUS5, uniqueCWR5, uniqueWEP6)

#bind the two df together 
sosSummary2 <- rbind(sosSummary2, nlgrpCollection)
rownames(sosSummary2)[3] <- "NLGRP Collection"
sosSummary2


```


# Uses

Report on distributions out of NPGS of materials from SOS, by year, by receiver, by species or taxonomic group
```{r}
deliveryData <-read.csv("H:\\SOS_Project\\analysisData\\sosOrdersTotal.csv")
print(deliveryData)

# select all rows that were delivered to ARS locations 
ars <- filter(deliveryData, category_code == "UARS")
ars <- ars[,1:2]
labelsColumns <- c('Number of Deliveries', "Reason for Delivery" )
values2 <- c("Replenishment", "Germination", "Observation"
                              ,"Information only", "Transfer", "Backup"
                              ,"Distribution")
ars$order_type_code <- values2

colnames(ars) <- labelsColumns
colnames(ars)
ars<- ars[order(-ars$'Number of Deliveries'),]

print(ars)


```

Investigate reasons for use- narrative description in GRIN Global
How many are CWR and impacts

# Discussion

- W6 doesnt have the resources to manage seeds indefinitely
- Constraints / challenges to adopting more wild plants
- Constraints with regard to getting materials into long term storage
- Challenge of not having seed testing techniques/rules worked out for species


## Quarterly Results {.tabset .tabset-fade .tabset-pills}
